WFH and broadband speed (title needs rework)

### 2020-12-29 paper/v1_old/broadband.speed.covid.tex
abstract + directories

\documentclass[Royal,times,sageh]{sagej}

@@

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{dcolumn}
\def\tightlist{}
\newcommand{\beginappendix}{\setcounter{table}{0} \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}

@@

\title{Working from home and digital divides: resilience in the time of a
pandemic}

@@

\author{A. Uthor*\affilnum{1,2}, O. Tro\affilnum{2}, O. Vriga\affilnum{3}}

\affiliation{\affilnum{1}{Department of Incredible Research, University A, City A, Country A}\\\affilnum{2}{Department of Applied Things, University B, City B, Country B}\\\affilnum{3}{Very Important Stuff Committee, Institute C, City C, Country C}}

\corrauth{Corresponding author name, This is sample corresponding address.}

\email{\href{mailto:correspondingauthor@protonmail.com}{\nolinkurl{correspondingauthor@protonmail.com}}}

@@

\keywords{covid; internet; working from home; broadband speed; time-series
clusters;}

@@

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds, may reinforce or redress the spatial
and social dimensions of the digital division exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. Our analysis enables
us to better understand how the spatial and social distribution of both
occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand. We will also
consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era, and broadband services and
infrastructure must be fit for purpose.

@@

In this analysis, the terms `telecommuting' and `working from home\^{}
{[}See also the popular social media hashtag \#WFM{]}' are used
interchangeably, as most remote labour during the Covid-19 crisis was
carried out in the homes of individual employees rather than any other
location \citep{eurofound2020}. However, it should be noted that
previous research has explored how telecommuting can occur in other
places, including satellite offices or on public transport
\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
also used a variety of definitions to measure the level of telecommuting
within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or
part-time, and those who use digital technologies to work remotely
full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. For example, studies from the United States, the
Netherlands, and the UK found that telecommuters are most likely to hold
professional, managerial, and technical occupations where the workforce
is better educated and wealthier, and that there is suppressed demand
among women and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.

@@

Furthermore, these multi-layered digital divides intersect with material
divides and the economic geography of the UK. Following the regional
economic resilience literature, which underlines the differentiated
capacity of cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}, different places have
different industrial and occupational profiles, and these affect the
aggregated potential capacity of places for telecommuting. Such profiles
are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide
\citep{martin_north_south}. Various studies have illustrated severe
inequalities between the north and the south regions of England
regarding, for example, skills and human capital, unemployment,
productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers some of the highest level
of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
digital divides associated, to a certain extent, with or shaped by the
geography of the UK, but the intersection of the digital and material
divides affects the capacity of places to overcome some of the economic
effects of the Covid-19 pandemic. Importantly, this is the first time
that digital technologies became an essential tool for economic
resilience for such a great part of the population.

@@

Based on the outcomes of the time-series clustering, we identify \(J\)
distinct and crisp clusters. We then regress this cluster membership
against a vector \(X_{i}\) of socio-economic and geographic variables,
which are discussed in detail in the relevant Section
\protect\hyperlink{sec:4.2}{4.2}, in order to explore how the different
patterns might support or undermine efforts to work from home and
maintain safe productivity and whether they reinforce existing spatial
and social inequalities. This analysis enables us to provide a more
nuanced understanding of how telecommuting and technology intersect at a
time of extreme demand, and what lessons this time has for a future
where telecommuting is likely to remain a common means of accessing work
and broadband services, as well as infrastructure, must be fit for
purpose.

Our analysis is not free of limitations as, for instance, although we do
have individual data we cannot identify individuals or households and,
therefore, we need to aggregate our data at the LAD level. However, such
individual data could only be found though surveys which by definition
result in limited samples and, importantly, less detailed data about the
quality of internet connectivity.

@@

The temporal profiles of the local authority clusters have been
summarised in Figures \ref{UpClusterL} and \ref{UpClusterS} and Table
\ref{up.cluster.descr}. The graphs show a composite profile of mean
upload speeds per hour per day for each cluster, with the largest, in
terms of the LAD membership and population, five clusters in Figure
\ref{UpClusterL}, and the next six in Figure \ref{UpClusterS}. These
figures and table provide a comprehensive overview of the quality and
reliability of experienced broadband in different parts of the UK.

@@

The second largest cluster, comprising \(126\) local authorities and
over \(20\) million people, is cluster \(13\), which has the slowest
aggregate mean upload speed of any of the clusters, and the second
highest ratio of the standard deviation to the mean. This suggests that
those living in local authorities in this cluster experienced some of
the lowest quality broadband services in terms of upload speeds in the
UK. However, the variation in upload speeds in cluster \(13\), which can
be an indication of its reliability, does not seem to disproportionately
affect the morning peak from \(9:00\)-\(10:59\), as upload speeds are,
on average, only \(2.6\)\% slower than in the evening peak period
between \(19:00\) and \(20:59\), when entertainment purposes are likely
to be using the most bandwidth. In comparison, the five LADs that are
home to \(343\) thousand people in cluster \(1\) not only experience the
second slowest mean upload speeds and the highest ratio of standard
deviation to the mean, but are also much more affected during the
morning peak.

Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
LADs home to \(29.5\) million people, experienced aggregate mean upload
speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
lower than the other three large clusters and most of the smaller
clusters, suggesting a middling quality of service. However, the
temporal profile for cluster \(6\) in Figure \ref{UpClusterL} shows that
upload speeds are highest at \(6:00\) on a Monday morning and peak
between \(23:00\) and midnight on Wednesday and Thursday, but tend to be
lower during the working day. Indeed, experienced mean upload speeds in
the morning peak are \(4.4\)\% lower than in the evening peak -- a
greater, more noticeable change than any of the other large clusters
experience, but smaller than any of the clusters with temporal profiles
shown in Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have
mean upload speeds under \(10\)Mb/s, but higher than clusters \(1\) and
\(13\), and their standard deviation is not dissimilar. However, this
masks great variation in when lower speeds are experienced, with the
mean upload speeds much lower between \(9:00-10:59\) than between
\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning in
cluster \(12\).

Indeed cluster \(12\) is the only cluster to experience higher speeds in
the evening peak, compared to the morning peak, suggesting that
widespread telecommuting has generally changed the temporal profile of
internet activity throughout the UK. Yet even if all but one cluster is
showing slower speeds in the morning than the evening, the reliability
of internet services in different clusters during the working day still
varies considerably. Interpreting this variation from the large spikes
and dips shown on Figures \ref{UpClusterL} and \ref{UpClusterS} is
difficult, but the statistics in Table \ref{up.cluster.descr} show that
clusters \(9\) and \(11\) have the most reliable internet services. The
ratio of standard deviation to mean in both these clusters is below
55\%, and the ratio of upload speeds in the peak periods indicates that
speeds are only about 2\% slower in the morning than the evening. Mean
speeds are also higher than in any other cluster, excluding cluster
\(2\), where measures of reliability suggests poorer performance.

@@

In summary, the local authorities in clusters \(9\) and \(11\)
experienced resilient broadband internet that could support high levels
of telecommuting. Those in clusters \(2\), \(3\), \(7\), and \(10\) also
experience higher than average mean speeds and rank middle to high on
measures of service reliability. These LADs are \emph{not} on the wrong
side of the first level digital divide, but how likely are they to be
able to take advantage of their resilient ICT infrastructure and
services? Meanwhile, cluster \(6\) is not only the largest in terms of
number of LADs and population, it has the closest mean upload speed to
the pre-clustered average for the whole sample. As well as average
quality internet services, those in cluster \(6\) also experience
average reliability for work purposes, ranking fifth behind the four
other clusters with populations over one million, but ahead of the
smaller clusters. Clusters \(8\) and \(12\) are also close to average
mean upload speeds, but show very different patterns in terms of
reliability, whilst clusters \(1\) and \(13\) appear to suffer most from
a lack of quality internet services, with slow speeds and high standard
deviations. With those in cluster \(1\) in particular more likely to
experience that poor reliability during the morning peak, is this first
level digital divide occurring in areas where few are occupationally
able to telecommute anyway, and what are the implications for economic
resilience?

@@

Using an auxiliary multinomial regression, we test whether the clusters
that have higher mean speeds and more reliable services consist of LADs
that are more urban, affluent, and / or more likely to benefit from a
choice of high quality internet services. We also estimate which of our
clusters are more likely to have a higher proportion of occupations
which enable telecommuting because of the nature of the work. The
results of the auxiliary regression are presented in Table \ref{aux}.
The dependent variable is the LAD cluster membership as described in the
\protect\hyperlink{sec:3}{methods and data} section and equation
\ref{eq1} and each column represent a different cluster. The reference
case is cluster \(4\), which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand
people. Mean, experienced upload speeds in cluster \(4\) are close to
both the average speeds for the \(13\) clusters (\(9.9\)Mb/s) and the
pre-clustered average for the whole sample (\(9.3\)Mb/s) as shown in
Table \ref{up.cluster.descr}. However, the standard deviation for
cluster \(4\) and the difference between average speeds in the morning
compared to the evening peak periods are indications of worse
reliability than many of the other clusters. Hence, the results in Table
\ref{aux} should be seen as relative rather than absolute probabilities.

First, we control for the number of speed tests run per cluster
inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
Virgin Media internet connections. Regarding the former, we expect
people in LADs with more unreliable connections to test more often their
internet speeds and the results validate our priors. Regarding the
former, fast Virgin Media cable connections have historically only been
available to \(45\)\% of premises in the UK \citep{ofcom2016}, where the
more lucrative and competitive market originally attracted the cable TV
provider. Those in clusters \(2\), \(3\), \(9\) and \(11\) benefit from
a higher proportion of Virgin connections, which is an indication that
people in these clusters are more likely to live in urban areas, with
more choice of broadband services, or in other words, are more likely to
be on the right side of the first level, infrastructure-based digital
divide.

Then, we employ distance from London and for the nearest metropolitan
area (including London) as two variables depicting peripherality, urban
structure and, potentially, first level digital divides. The broadband
speed tests run in the authorities in cluster \(3\) are more likely to
be taking place close to London than those run in any of the other
clusters, a result that makes sense considering two of the four
authorities in cluster \(3\) are the London commuter towns of Harlow and
Luton. However, even though London was also included in the variable
calculating distance from the centre of one of either the ten largest
metropolitan areas in England, or Glasgow or Cardiff, tests run in
cluster \(3\) are likely to be furthest away. Still, whilst significant,
the effect is tiny and the other members of cluster \(3\) are Corby and
Eastbourne, two towns which, whilst not part of metropolitan areas, are
also not in remote parts of the country, and are home to an estimated
population of \(175\) thousand people between them.

\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-6}
\end{figure}

Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
of a large metropolitan area, even though the four local authorities of
cluster \(7\) include no central urban boroughs and only one LAD that is
part of a metropolitan area of governance - Tameside in Greater
Manchester. This may explain why those in cluster \(7\) are not served
by Virgin Media. It also is a demonstration of the complexity of
experienced broadband upload speeds as captured by time-series
clustering, and their likely interaction with first level digital
divides. Cluster \(1\), which our analysis suggests lacks broadband
resilience, contains five, mainly rural authorities. However, they are
scattered around the country -- Ceredigion in West Wales, Darlington in
the Northeast, Eden between the Lake District and the North Penines,
Rossendale in the Northwest, and Rutland in the East Midlands -- and
therefore, the results in Table \ref{aux} indicate that these
authorities are not as far from the centre of a metropolitan area as
authorities in cluster \(3\), but furthest from London.

Internet resilience, as measured here, seems to be more nuanced than our
-- arguably crude -- dummy variable depicting the North-South economic
divide either, which assigns \(1\) to LADs locates in Greater London,
Southeast, Southwest and East of England regions, and \(0\) to the rest.
The authorities most likely to be in the South are those in cluster
\(13\), which was identified as having the slowest mean upload speeds of
any of the clusters, and thus a low level of service. However, cluster
\(13\) does include some rural, remote areas of the country, such as
Northwest Scotland, Cornwall and Powys in Wales as shown on Figure
\ref{map.up.clusters}. It also includes the major metropolitan centres
of Bristol, Liverpool, and Newcastle, as well as nine (of \(32\)) London
Boroughs, including the City of London and plenty of home county and
suburban areas. Although by the standard deviation measure cluster
\(13\) has unreliable internet services, speed variation was less during
the morning peak, suggesting that the estimates for reliability are
inconsistent. Considering that this is one of the largest clusters, and
thus the averages incorporate more noise than some of the smaller
clusters, it may be that the LADs in this cluster do not all suffer
equally from a first level digital divide.

Yet we need to consider the results for other variables in Table
\ref{aux} to better determine whether clusters \(1\) and \(13\), which
appear to suffer most from a lack of quality internet services, are also
more likely to have a low skilled workforce, less able to benefit from
telecommuting. Cluster \(1\) has the lowest propotion of educated people
of any of the clusters, and the lowest earnings, despite recording the
highest proportion of managerial, professional jobs, and the second
highest proportion of tech jobs. Rural areas such as those in cluster
\(1\) are often home to many retired people, which might explain these
results or perhaps, as these figures are relative, we could note that
cluster \(1\) also has a greater proportion of skilled trades than other
clusters. Either way, it seems likely that the first level digital
divide reinforces other inequities in cluster \(1\). In comparison,
those in cluster \(13\) are more likely to earn more and have a better
education. Cluster \(13\) also has the most businesses per inhabitant,
but is somewhere in the middle in terms of job density. If this is an
indication of a high number of SMEs, it can explain the variable
internet quality, considering that small businesses have not been seen
as the most valuable customers for higher speed broadband services
\textbf{insert ref}.

Cluster \(6\) is the largest cluster, and thus, like cluster \(13\),
there is more noise within the averages we use to measure internet
quality and reliability, and the result indicates an average level of
service, even if better than LADs in cluster \(13\). Cluster \(6\) also
falls towards the middle of the clusters on many of the socio- economic
variables in Table \ref{aux}. It ranks third or fourth out of the eleven
clusters in the table in terms of the likelihood of having a higher
proportion of managerial, professional, and tech jobs, as well as higher
earnings, but is fourth from bottom in terms of education. The LADs in
cluster \(6\) are also diverse, with few truly remote areas, but many
urban areas in both the North, including Birmingham, Leeds and
Sheffield, and South, including twelve London Boroughs, and many
suburban areas and smaller cities like Oxford and Cambridge. The capital
cities of the other UK nations, Belfast, Cardiff and Edinburgh, are also
in this cluster, suggesting perhaps that the lower level of reliability
discussed in Section \protect\hyperlink{sec:4.1}{4.1} may be as a result
of increased demand, e.g.~for telecommuting.

In comparison, clusters \(9\) and \(11\), which enjoy the most resilient
internet connections, are characterised by higher share of occupations
that may benefit from the ability to telecommute (e.g.~managerial).
Also, LADs in Cluster \(9\) are more likely to host highly educated
individuals with higher earning than cluster \(11\), which has fewer
businesses per inhabitant and lower job density. Indeed, cluster \(11\)
has a negative coefficient for the NVQ4 variable, whilst for cluster
\(9\) the coefficient is positive, and cluster \(11\) has the second
highest job density and cluster \(9\) the second lowest. The former
might be an indication of the higher returns that resilient broadband
infrastructure can generate for cluster \(9\) versus cluster \(11\).
Indeed, we did detect slightly more slowdown in the AM peak in cluster
\(9\) than in cluster \(11\) (see Table \ref{up.cluster.descr}).
Meanwhile, a propensity for a higher job density indicate higher
urbanization rate in LADs in cluster \(11\) than those in cluster \(9\).

LADs in cluster \(9\) are also less likely to be in the South and those
in cluster \(11\) more likely, although the coefficient for cluster
\(9\) might be skewed by the presence of the Scottish cities of Glasgow
and Aberdeen, as Scotland has a different economic profile than England.
Still, both clusters consist of larger LADs in terms of population,
including the urban centre of Manchester and three of the nine other
districts of Greater Manchester, Nottingham, both Portsmouth and
Southampton, eight London Boroughs, and four of the seven constituent
authorities of the West Midlands Combined Authority. Among the \(65\)
LADs are also a number of other tightly bounded urban areas, such as
Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at
the centre of less confined districts like Burnley, the digital hub of
Milton Keynes, and Northampton. Thus, our findings confirm the
importance of geography and urban locations for being on the right side
of the first level digital divide, and suggest that internet resilience
cuts across other digital and economic divides.

>>>>

\documentclass[,]{sagej}

@@

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{dcolumn}
\def\tightlist{}

@@

\title{WFH and broadband speed (title needs rework)}

@@

\author{A. N. Author\affilnum{}, John Smith\affilnum{}}

\affiliation{}

@@

\keywords{covid; internet; working from home; broadband speed; time-series
clusters}

@@

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds, may reinforce or redress the spatial
and social dimensions of the digital division exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. Our analysis enables
us to better understand how the spatial and social distribution of both
occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand. We will also
consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era, and broadband services and
infrastructure must be fit for purpose. \textbf{LET'S LEAVE IT FOR NOW,
BUT I THINK WE CAN CRYSTALISE MORE THE RQ}

@@

In this analysis, the terms `telecommuting' and `working from home' are
used interchangeably, as most remote labour during the Covid-19 crisis
was carried out in the homes of individual employees rather than any
other location \citep{eurofound2020}. However, it should be noted that
previous research has explored how telecommuting can occur in other
places, including satellite offices or on public transport
\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
also used a variety of definitions to measure the level of telecommuting
within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or
part-time, and those who use digital technologies to work remotely
full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. For example, studies from the United States, the
Netherlands, and the UK found that telecommuters are most likely to hold
professional, managerial, and technical occupations where the workforce
is better educated and wealthier, and that there is suppressed demand
among women and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.

@@

Furthermore, these multi-layered digital divides intersect with material
divides and the economic geography of the UK. Following the regional
economic resilience literature, which underlines the differentiated
capacity of cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}, different places have
different industrial and occupational profiles, and these affect the
aggregated potential capacity of places for telecommuting. Such profiles
are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide
\citep{martin_north_south}. Various studies have illustrated severe
inequalities between the north and the south regions of England
\textbf{replace `UK'} regarding, for example, skills and human capital,
unemployment, productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers some of the highest level
of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
digital divides associated, to a certain extent, with or shaped by the
geography of the UK, but the intersection of the digital and material
divides affects the capacity of places to overcome some of the economic
effects of the Covid-19 pandemic. Importantly, this is the first time
that digital technologies became an essential tool for economic
resilience for such a great part of the population.

@@

Based on the outcomes of the time-series clustering, we identify \(J\)
distinct and crisp clusters. We then regress this cluster membership
against a vector \(X_{i}\) of socio-economic and geographic variables,
which are discussed in detail in the relevant Section
{[}4.2{]}{]}(\#sec:4.2), in order to explore how the different patterns
might support or undermine efforts to work from home and maintain safe
productivity and whether they reinforce existing spatial and social
inequalities. This analysis enables us to provide a more nuanced
understanding of how telecommuting and technology intersect at a time of
extreme demand, and what lessons this time has for a future where
telecommuting is likely to remain a common means of accessing work and
broadband services, as well as infrastructure, must be fit for purpose.

@@

The temporal profiles of the local authority clusters have been
summarised in Figures \ref{UpClusterL} and \ref{UpClusterS} and Table
\ref{up.cluster.descr}. The graphs show a composite profile of mean
upload speeds per hour per day for each cluster, with the largest, in
terms of the LAD membership and population, five clusters in Figure
\ref{UpClusterL}, and the next six in Figure \ref{UpClusterS}. These
figures and table provide a comprehensive overview of the quality and
reliability of experienced broadband in different parts of the UK, the
temporal clusters offering a novel approach to understanding spatial
disparity.

@@

The second largest cluster, comprising \(126\) local authorities and
over \(20\) million people, is cluster \(13\). Cluster \(13\) has the
slowest aggregate mean upload speed of any of the clusters, and the
second highest ratio of the standard deviation to the mean. This
suggests that those living in local authorities in this cluster
experienced some of the lowest quality broadband services in terms of
upload speeds in the UK. However, the variation in upload speeds in
cluster \(13\), which can be an indication of its reliability, does not
seem to disproportionately affect the morning peak from
\(9:00\)-\(10:59\), as upload speeds are, on average, only \(2.6\)\%
slower than in the evening peak period between \(19:00\) and \(20:59\),
when entertainment purposes are likely to be using the most bandwidth.
In comparison, the five LADs that are home to \(343\) thousand people in
cluster \(1\) not only experience the second slowest mean upload speeds
and the highest ratio of standard deviation to the mean, but are also
much more affected during the morning peak.

Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
LADs home to \(29.5\) million people, experience aggregate mean upload
speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
lower than the other three large clusters and most of the smaller
clusters, suggesting a middling quality of service. However, the time
profile for cluster \(6\) in Figure \ref{UpClusterL} shows that upload
speeds drop quickly from \(6:00\) to \(7:00\) on a Monday and peak
between \(23:00\) and midnight on Wednesday and Thursday, but tend to be
lower during the working day. Indeed, experienced mean upload speeds in
the morning peak are \(4.4\)\% lower than in the evening peak -- a
greater, more noticeable change than any of the other large clusters
experience, but smaller than any of the clusters with temporal profiles
shown in Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have
mean upload speeds under \(10\)Mb/s, but higher than clusters \(1\) and
\(13\), and their standard deviation is not dissimilar. However, this
masks great variation in when lower speeds are experienced, with the
mean upload speeds much lower between \(9:00-10:59\) than between
\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning in
cluster \(12\).

\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-5}
\end{figure}

Indeed cluster \(12\) is the only cluster to experience higher speeds in
the evening peak, compared to the morning peak, suggesting that
widespread telecommuting has generally changed the temporal profile of
internet activity throughout the UK. Yet even if all but one cluster is
showing slower speeds in the morning than the evening, the reliability
of internet services in different clusters during the working day still
varies considerably. Interpreting this variation from the large spikes
and dips shown on Figures \ref{UpClusterL} and \ref{UpClusterS} is
difficult, but the statistics in Table \ref{up.cluster.descr} show that
clusters \(9\) and \(11\) have the most reliable internet services. The
ratio of standard deviation to mean in both these clusters is below
55\%, and the ratio of upload speeds in the peak periods indicates that
speeds are only about 2\% slower in the morning. Mean speeds are also
higher than in any other cluster, excluding cluster \(2\), where
measures of reliability suggests poorer performance.

@@

In summary, the local authorities in clusters \(9\) and \(11\)
experienced resilient broadband internet that could support high levels
of telecommuting. Those in clusters \(2\), \(3\), \(7\), and \(10\) also
experience higher than average mean speeds and rank high to middle on
measures of service reliability. These LADs are \emph{not} on the wrong
side of the first level digital divide, but how likely are they to be
able to take advantage of their resilient ICT infrastructure and
services? Meanwhile, cluster \(6\) is not only the largest in terms of
number of LADs and population, it has the closest mean upload speed to
the pre-clustered average for the whole sample. As well as average
quality internet services, those in cluster \(6\) also experience
average reliability for work purposes, ranking fifth behind the four
other clusters with populations over one million, but ahead of the
smaller clusters. Clusters \(8\) and \(12\) are also close to average
mean upload speeds, but show very different patterns in terms of
reliability, whilst clusters \(1\) and \(13\) appear to suffer most from
a lack of quality internet services, with slow speeds and high standard
deviations. With those in cluster \(1\) in particular more likely to
experience that poor reliability during the morning peak, is this first
level digital divide occurring in areas where few are occupationally
able to telecommute anyway, and what are the implications for economic
resilience?

@@

Using auxiliary regressions, we test whether the clusters that have
higher mean speeds and more reliable services consist of LADs that are
more urban, affluent, and / or more likely to benefit from a choice of
high quality internet services. We also estimate which of our clusters
are more likely to have a higher proportion of occupations which enable
telecommuting because of the nature of the work. The results of these
auxiliary regressions are presented in Table \ref{aux}. The dependent
variable is the LAD cluster membership as described in the
\protect\hyperlink{sec:3}{methods and data} section and equation
\ref{eq1} and each column represent a different cluster. The reference
case is cluster \(4\), which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand
people. Mean, experienced upload speeds in cluster \(4\) are close to
both the average speeds for the \(13\) clusters (\(9.9\)Mb/s) and the
pre-clustered average for the whole sample (\(9.3\)Mb/s) as shown in
\ref{up.cluster.descr}. However, the standard deviation for cluster
\(4\) and the difference between average speeds in the morning compared
to the evening peak periods are indications of worse reliability than
many of the other clusters. Hence, the results in Table \ref{aux} should
be seen as relative rather than absolute probabilities.

First, the number of speed tests run per cluster inhabitant between
\(9:00\)-\(10:59\) is an indication of satisfaction or at least a lack
of concern over broadband speeds and quality of service. People in LADs
assigned to clusters \(6\) and \(13\) were more likely to ran speed
tests (at a per capita basis) at this time of day, whilst people in
cluster \(11\) were the least likely ones. This high level of
satisfaction matches the high quality and reliability of service we
would expect from the analysis in Section {[}4.1{]}{]}(\#sec:4.1). Those
in clusters \(2\), \(3\), \(9\) and \(10\) were also less likely to ran
speed tests per capita than those in the other six multi-authority
clusters. Cluster \(7\) was the only cluster with higher mean speeds and
middle to high measures of reliability which was more likely to run more
speed tests per capita. This might be a reflection of the choice of ISP
and connection options in those LADs, as, along with cluster \(12\),
those in cluster \(7\) were least likely to have Virgin Media
connections. Fast Virgin Media cable connections have historically only
been available to \(45\)\% of premises in the UK \citep{ofcom2016},
where the more lucrative and competitive market originally attracted the
cable TV provider. Those in clusters \(2\), \(3\), \(9\) and \(11\)
benefit from a higher proportion of Virgin connections, which is an
indication that people in these clusters are more likely to live in
urban areas, with more choice of broadband services, or in other words,
are more likely to be on the right side of the first level,
infrastructure-based digital divide.

We employ distance from London and for the nearest metropolitan area
(including London) as two more variables depicting peripherality, urban
structure and, potentially, first level digital divides. The broadband
speed tests run in the authorities in cluster \(3\) are more likely to
be taking place close to London than those run in any of the other
clusters, a result that makes sense considering two of the four
authorities in cluster \(3\) are the London commuter towns of Harlow and
Luton. However, even though London was also included in the variable
calculating distance from the centre of one of either the ten largest
metropolitan areas in England, Glasgow or Cardiff, tests run in cluster
\(3\) are likely to be furthest away. Still, whilst significant, the
effect is tiny and the other members of cluster \(3\) are Corby and
Eastbourne, two towns which, whilst not part of metropolitan areas, are
also not in remote parts of the country, and are home to an estimated
population of \(175\) thousand people between them. \textbf{move maps to
here?} Meanwhile, LADs in cluster \(7\) are most likely to be near the
centre of a large metropolitan area, even though the four local
authorities of cluster \(7\) include no central urban boroughs, and only
one LAD that is part of a metropolitan area of governance - Tameside in
Greater Manchester. This may be why those in cluster \(7\) are not
served by Virgin Media. It also is a demonstration of the complexity of
experienced broadband upload speeds as captured by time-series
clustering, and their likely interaction with first level digital
divides. Cluster \(1\), which our analysis suggests lacks broadband
resilience, contains five, mainly rural authorities. However, they are
scattered around the country -- Ceredigion in West Wales, Darlington in
the Northeast, Eden between the Lake District and the North Penines,
Rossendale in the Northwest, and Rutland in the East Midlands -- and
therefore, the results in Table \ref{aux} indicate that these
authorities are not as far from the centre of a metropolitan area as
authorities in cluster \(3\), but furthest from London.

Internet resilience, as measured here, seems to be more nuanced than our
-- arguably crude -- dummy variable depicting the North-South economic
divide either, which assigns \(1\) to LADs locates in Greater London,
Southeast, Southwest and East of England regions. The authorities most
likely to be in the South are those in cluster \(13\), which was
identified as having the slowest mean upload speeds of any of the
clusters, and thus a low level of service. However, cluster \(13\) does
include some rural, remote areas of the country, such as Northwest
Scotland, Cornwall and Powys in Wales as shown on Figure
\ref{map.up.clusters}. It also includes the major metropolitan centres
of Bristol, Liverpool, and Newcastle, as well as nine (of \(32\)) London
Boroughs and plenty of home county and suburban areas. Although by the
standard deviation measure cluster \(13\) has unreliable internet
services, speed variation was less during the morning peak. More
broadly, the estimates for this variable are not consistent between the
two clubs of LAD clusters: those who are in the wrong and the right side
of digital divides.

Regarding the rest of the socio-economic variables we see a varied
picture. Clusters \(9\) and \(11\), which are the ones benefiting by the
most resilient internet connections are likely to consist of LADs with
higher share of managerial and highly skilled occupations as well as
higher average earnings. Interestingly, both clusters are characterised
by lower share of occupations that are not usually associated with
teleworking (professional, associated professional and technical,
administrative, leisure and machine operation jobs), but also by lower
number of business per habitant and. However, the likelihood of hosting
highly skilled residents is not consistent with cluster \(11\) having a
negative coefficient for the NVQ4 variable. This might be an indication
of the higher returns that resilient broadband infrastructure generates
for cluster \(9\) versus cluster \(11\). While both of these cluster
consist more likely by large LADs in terms of population, job density if
higher for LADs in cluster \(11\) indicating more labour intensive
economic structures contrary to LADs in cluster \(9\).

A similar divided picture can be built for the next tier of clusters:
\(2\), \(3\), \(7\), \(10\). LADs in these clusters experience higher
than average mean speeds and rank high to middle on measures of service
reliability. As mentioned before, LADs in these clusters are on the
right side of the first level digital divide. However, while all of
them are more likely to be populated by people employed in managerial
jobs, only LADs in cluster \(2\) and \(10\) are more likely to have
highly skilled work force, which can more easily take advantage of
working from home opportunities. A more mixed picture can be built for
LADs in clusters \(3\) and \(7\), which are associated, on the one hand,
with lower skills and, on the other hand, with higher likelihood of
individuals in managerial occupations. A similar contradicting pattern
can be illustrated for cluster \(6\), which is the largest one with
average quality of internet service. Arguably, LADs in clusters \(2\)
and \(10\) are the ones benefiting more by resilient internet
infrastructure in an era that working from home became a vehicle for
economic resilience.

Moving on to the next tier of clusters \(8\) and \(12\), which are
characterised by LADs enjoying close to average mean upload speeds, but
different patterns of internet reliability, we see a differentiated
picture. On the one hand, both of them are likely to host individuals
employed in managerial occupations while the opposite applies on
skilled, professional and leisure occupations. On the other hand, LADs
in cluster \(8\) are characterised by a very small positive and LADs in
cluster \(12\) by a negative likelihood to host highly skilled
individuals. Moreover, LADs in these two clusters have a rather similar
profile regarding the occupational variables, but different tendencies
regarding the stock of businesses. In total, LADs in these clusters do
not benefit by great internet services and could only marginally benefit
by working from home opportunities.

Lastly, clusters \(1\) and \(13\), which appear to suffer most from a
lack of quality internet services are more likely to also suffer from
low skilled workforce. LADs in cluster \(13\). Contradictory, the
likelihood of workforce employed in managerial occupations is positive
for LADs in both clusters, but only individuals in LADs in cluster
\(13\) are more likely to enjoy high earnings.

\protect\hyperlink{sec:appendix}{Appendix}
